The importance of a complete statistical description of dependence between variables for ecology and related fields

### 2018-09-05 Paper.Rmd
modified

Though a few examples have been found which shows copula approach to model dependencies in bioinformatics [@Kim2008], medical research [@Emura2016] or a short period study of intersite dependence of rainfall data [@Serinaldi2008], the multivariate copula approach still, we believe, remains unexplored in ecology or related field and thus have potential scope to better capture the evolution of ecological time series data. So, in ecological perspective, when we find a dominantly lower (upper) tail dependence structure, it means both incidence are rare (abundant) synchronously. If any one of them be environmental variable and the other is an observable (like abundance of a species) then it can also explain the causal relationship between them. Here, we analyzed the dependence structure for several bivariate and multivariate ecological, community and biogeochemical datasets. In addition to a copula based \nameref{Model_selection}, we also carried out a \nameref{NPA} with three different statistics to verify our model selection based results.

@@

Q1/2: How are copula structure of ecologically influential climate variables changing as part of climate change? We now know the copula structure of environmental variables can induce (sometimes similar but not the same) copula structure in the pops, but we don’t really know how a change in the environment cop structure would make what kind of change in the pop cop structure. How?

@@

We show here that if a non-normal copula structure describes the dependence of an environmental
variable between two locations, then there can be a non-normal copula structure for the
dependence of populations in the locations that are influenced by the environmental variables.

@@

To test that our approach is of two way : first, we calculated the parameters of the noise-copula (for example, we chose Clayton with lower tail dependence and Survival Clayton with upper tail dependence here) using `iRho` or `iTau` function from `copula` package for Spearman's Rho or Kendall's Tau, respectively. Then we simulated our model (\ref{eq.AR1}) and estimated parameters of the outcome which was a bivariate population-copula, with last $500$ simulated points for each habitat patch (see Fig. \ref{SM-fig_Cause4copula_params_GOF_500}). We also tested the goodness of fit of the estimated parameters for population-copula and its standard error due to sample size and reported p-values for both statistics (Cramer-von Mises and Kolmogorov-Smirnov). p-values greater or equal to $0.05$ (above the dashed line in Fig. \ref{SM-fig_Cause4copula_params_GOF_500}) were considered as a good fit of estimated parameters. Fig. \ref{SM-fig_Cause4copula_params_GOF_2500} represents similar figure as of Fig. \ref{SM-fig_Cause4copula_params_GOF_500} but here we considered last $2500$ simulated population data points for our parameters-estimation, thus making our comparison more sensitive on basis of p-value threshold at $0.05$. (For code see [`Plotter_Cause4copula_GOF`](https://github.com/sghosh89/BIVAN/blob/master/Cause4copula.R) function).

@@

We simulated our model (\ref{eq.AS}) for different values of $b$ with $a=0.2$, $r=0.8$. $10000$ points were generated from each local and global fluctuations to apply on the initial populations and then after simulation we kept only last $1000$ time step of simulation to plot the population-copula for a specific $b$ and a specific functional form of $g$  (For code see [`Simulator_asym_sens`](https://github.com/sghosh89/BIVAN/blob/master/asym_sensitivity.R) function). Fig. \ref{SM-fig_ms_asym_b_0.1} and Fig. \ref{SM-fig_ms_asym_b_0.9} show these results for $b=\pm0.1$ and $b=\pm0.9$, respectively. We also carried out model selection and non-parametric statistical analysis with each copula of those figures and summarized our results in Tables \ref{SM-tab_summary_stat_asym_sens} - \ref{SM-tab_stat_npa_asym_8}. Table \ref{SM-tab_summary_stat_asym_sens} represent the model selection results with AIC weights for best fitted copula and normal copula, model averaged lower and upper tail dependence, p-values (both Cramer-von Mises statistic and Kolmogorov-Smirnov statistic) for goodness of fit test. In each case, successful bootstrap was $100\%$. According to our model selection approach, for $b=\pm0.1$, non-normal copula (with higher AIC weights) were always best-fit with some asymetric tail dependence whereas for $b\pm0.9$, most of the time normal was best. It is probably because the populations get long-time exposure to environmental changes for slower growth(or decay) rate and so global fluctuation can affect them more efficiently. Our non-parametric analysis results (Tables \ref{SM-tab_stat_npa_asym_1} -\ref{SM-tab_stat_npa_asym_8}) are consistent with this model selection approach.

@@

@@

Here, we claim that because of significant tail dependence of the multivariate copula structure, the skewness of the time-series data averaged over all selective locations changed significantly from the individual skewness of time-series data taken at each pairwise location. Spatial mean of the time series data extracts that tail dependence of the multivariate data set, if any, and reflected in the averaged skewness. We tested this with null hypothesis $H_{0}$. If $H_{0}$ is true that means our multivariate copula actually came from normal copula structure with no tail dependence and thus should have the same skewness as of normal copula. So, first we generated $10000$ number of surrogate normal copula from our given multivariate copula keeping the marginal distribution and the correlation (spearman or kendall) same. Then we plotted the distribution of spatial mean of each surrogate time series data set and the given multivariate data. If $p$-value is less than $0.05$, then we reject the null hypothesis in support of significant tail dependence.

To verify that we used aphid-count data, aphid-firstflight data, plankton data and methane-flux data. For aphid data we choose only those locations for which atleast $30$ number of data points present in each time series when taken as pairwise locations. For plankton-UK-sea data and methane-flux data that threshold on number of data points are $45$ and $50$, respectively.

- Aphid count \label{skewness_count}

Figure \ref{SM-fig_skewness_count} shows that averaged skewness is much lower than expected if the aphid count data came from a normal copula. Here, $p$ value indicates the fraction of surrogate-skewness which are lower than the averaged skewness of the given data (red line in Figure \ref{SM-fig_skewness_count}). As $p<0.05$, we reject $H_{0}$. This result is consistent with the non-parametric analysis and model fitting results which indicated lower tail dependence.

- Aphid first flight \label{skewness_ff}

Figure \ref{SM-fig_skewness_ff} shows that averaged skewness is much higher than expected if the aphid first flight data came from a normal copula. Here, $p$ value indicates the fraction of surrogate-skewness which are greater than the averaged skewness of the given data (red line in Figure \ref{SM-fig_skewness_ff}). As $p<0.05$, we reject $H_{0}$ in support for upper tail dependence.

- UK Sea plankton \label{skewness_pns}

Figure \ref{SM-fig_skewness_pns} shows that averaged skewness is lower than expected if the plankton data came from a normal copula. Our previous results on non-parametric analysis and model fitting showed lower tail dependence. Here, as $p>0.05$, we cannot reject $H_{0}$, .......???

- Methane-flux data \label{skewness_methane}

Figure \ref{SM-fig_skewness_methane} shows that averaged skewness is lower than expected if the plankton north sea data came from a normal copula. Our previous results on non-parametric analysis and model fitting showed lower tail dependence. As $p<0.05$, we reject $H_{0}$. This result is consistent with the non-parametric analysis and model fitting results which indicated lower tail dependence.

@@

In order to study the effects of non-Normal copula structure on extinction risk for population of a species at different locations with different dispersal rules, we examined the following generalized multiple patch model where local population dynamics at each patch were governed by the Lewton-Cohen model ($r=0$) or Ricker model ($r \neq 0$) as the case may be;
<!--did I write correct spelling for lewton model?-->

@@

$r$ is the liner growth rate and $K$ is the carrying capacity of the environment. $\epsilon$ is the environmental noise which may take strong right tail (or left tail) dependence. We then considered each population $P(t+1)$ was updated depending upon $D$, the dispersal matrix. $D$ can take either forms; one case when each population patch on a linear chain interacted with its two nearest neighbors (we called it local dispersion) or when any patch can interact with the rest of all (we called it as global dispersion). We simulated Eq. \ref{eq.extrisk_model} for $10000$ simulations and calculated extinction risk after $25$ years with different possibilities (for code see [mydispersal.R](https://github.com/sghosh89/BIVAN/blob/master/mydispersal.R)) and summarized our findings as follows;

@@

    * For LC model, as expected, when extinction risk calculated after $25$ years, was plotted against each varying $d$ (=degree of dispersion), we found that extinction risk was higher for data came from left tail dependent environmental fluctuations than right tail dependent one. [see first plot of Fig. \ref{SM-fig_risk_local_disp} and Fig. \ref{SM-fig_risk_global_disp}]

@@

### Effects on Taylor's law

Spatial Taylor's law, which shows linear relationship between `log`(variance) and `log`(mean) of population densities of a species taken from different locations, is a well-known pattern found in ecology and related fields. Here, we want to examine the nature of the slope and how it changes if population time series (taken as $50$ years) from different locations (say, $numloc=25$) have some certain type of copula structure (namely, \textbf{C}, \textbf{F} and \textbf{SC}) other than the $\textbf{N}$ormal copula. First, we simulated (N $\times$ n $\times$ ns) dimensional normal copula with covariance $\rho =0.7$, where $(N=50,n=25,ns=10)$ are the length of timeseries, number of sampling locations and number of replica for that copula, respectively. Then we generated Kendall-correlation preserving \textbf{C}layton, \textbf{F}rank and $\textbf{S}$urvival \textbf{C}layton surrogates with same dimension of \textbf{N}ormal. With these four different types of copula we computed `log`(spatial variance) and `log`(mean) and plotted in Fig.\ref{fig_spTL}(A-D). We also tested statistically that whether linearity, homoskedasticity is maintained in each copula dataset or not and plotted our results alongwith root mean square error, intercept and slope in Fig.\ref{fig_spTL}(E-I) (For code see [`copsurrognd.R`](https://github.com/sghosh89/BIVAN/blob/master/copsurrognd.R) and [`TaylorsLawTools.R`](https://github.com/sghosh89/BIVAN/blob/master/TaylorsLawTools.R)).

>>>>

Though a few examples have been found which shows copula approach to model dependencies in bioinformatics [@Kim2008], medical research [@Emura2016], a short period study of intersite dependence of rainfall data [@Serinaldi2008], a recent study on characterizing and assessing risks in drought management [@She2018] or monitoring changes in precipitation-extremes [@Goswami2018], the multivariate copula approach still, we believe, remains unexplored in ecology or related field and thus have potential scope to better capture the evolution of ecological time series data. So, in ecological perspective, when we find a dominantly lower (upper) tail dependence structure, it means both incidence are rare (abundant) synchronously. If any one of them be environmental variable and the other is an observable (like abundance of a species) then it can also explain the causal relationship between them. Here, we analyzed the dependence structure for several bivariate and multivariate ecological, community and biogeochemical datasets. In addition to a copula based \nameref{Model_selection}, we also carried out a \nameref{NPA} with three different statistics to verify our model selection based results.

@@

This section is dedicated to explore question \hyperref[Intro]{\textbf{(Q2)}} from \nameref{Intro}. We want to find how are copula structure of ecologically influential climate variables changing as part of climate change? We now know the copula structure of environmental variables can induce (sometimes similar but not the same) copula structure in the populations, but we don’t really know how a change in the environment copula structure would make what kind of change in the population copula structure.

@@

Our hypothesis is if a non-normal copula structure describes the dependence of an environmental
variable between two locations, then there can be a non-normal copula structure for the
dependence of populations in the locations that are influenced by the environmental variables. This assumption is inspired by the the idea of Moran-effect [@Moran1953] which was originally used to explain synchronization of widely dispersed populations.

@@

To test that our approach is of two way : first, we calculated the parameters of the noise-copula (for example, we chose Clayton with lower tail dependence and Survival Clayton with upper tail dependence here) using `iRho` or `iTau` function from `copula` package for Spearman's Rho or Kendall's Tau, respectively. Then we simulated our model (\ref{eq.AR1}) and estimated parameters of the outcome which was a bivariate population-copula, with last $500$ simulated points for each habitat patch (see Fig. \ref{SM-fig_Cause4copula_params_GOF_500}). We also tested the goodness of fit of the estimated parameters for population-copula and its standard error due to sample size and reported p-values for both statistics (Cramer-von Mises and Kolmogorov-Smirnov). p-values greater or equal to $0.05$ (above the dashed line in Fig. \ref{SM-fig_Cause4copula_params_GOF_500}) were considered as a good fit of estimated parameters. Fig. \ref{SM-fig_Cause4copula_params_GOF_2500} represents similar feature as of Fig. \ref{SM-fig_Cause4copula_params_GOF_500} but here we considered last $2500$ simulated population data points for our parameters-estimation, thus making our comparison more sensitive on basis of p-value threshold at $0.05$. (For code see [`Plotter_Cause4copula_GOF`](https://github.com/sghosh89/BIVAN/blob/master/Cause4copula.R) function).

@@

We simulated our model (\ref{eq.AS}) for different values of $b$ with $a=0.2$, $r=0.8$. $10000$ points were generated from each local and global fluctuations to apply on the initial populations and then after simulation we kept only last $1000$ time step of simulated points to plot the population-copula for a specific $b$ and a specific functional form of $g$  (For code see [`Simulator_asym_sens`](https://github.com/sghosh89/BIVAN/blob/master/asym_sensitivity.R) function). Fig. \ref{SM-fig_ms_asym_b_0.1} and Fig. \ref{SM-fig_ms_asym_b_0.9} show these results for $b=\pm0.1$ and $b=\pm0.9$, respectively. We also carried out model selection and non-parametric statistical analysis with each copula of those figures and summarized our results in Tables \ref{SM-tab_summary_stat_asym_sens} - \ref{SM-tab_stat_npa_asym_8}. Table \ref{SM-tab_summary_stat_asym_sens} represent the model selection results with AIC weights for best fitted copula and normal copula, model averaged lower and upper tail dependence, p-values (both Cramer-von Mises statistic and Kolmogorov-Smirnov statistic) for goodness of fit test. In each case, successful bootstrap was $100\%$. According to our model selection approach, for $b=\pm0.1$, non-normal copula (with higher AIC weights) were always best-fit with some asymetric tail dependence whereas for $b\pm0.9$, most of the time normal was best. It is probably because the populations get long-time exposure to environmental changes for slower growth(or decay) rate and so global fluctuation can affect them more efficiently. Our non-parametric analysis results (Tables \ref{SM-tab_stat_npa_asym_1} -\ref{SM-tab_stat_npa_asym_8}) are consistent with this model selection approach.

@@

This section is attributed to the analysis of question \hyperref[Intro]{\textbf{(Q3)}} from \nameref{Intro}. Here we summarized our findings as follows;

@@

We hypothesized that because of distinct tail dependence of the multivariate copula structure, the skewness of the time-series data averaged over all selective locations changed significantly from the individual skewness of time-series data taken for each pairwise location. Spatial mean of the time series data extracted that tail dependence of the multivariate data set, if any, and reflected in the averaged skewness. We tested null hypothesis $H_{0}$ which presumed that multivariate copula actually came from normal copula structure with no tail dependence and thus should have the same skewness as of normal copula. So, first we generated $10000$ number of surrogate normal copula from a given multivariate copula keeping the marginal distribution and the correlation (spearman or kendall) same. Then we plotted the distribution of spatial mean of each surrogate time series data and of the given multivariate data. If $p$-value is less than $0.05$, then we reject the null hypothesis in support of significant tail dependence (and henceforth significantly different skewness).

To verify that we used our four multivariate datasets : aphid-count data, aphid-firstflight data, plankton data and methane-flux data. Figure \ref{SM-fig_skewness_count} showed that averaged skewness was much lower than expected if the aphid count data came from a normal copula. In that example, $p$ value was denoted by the fraction of surrogate-skewness which were lower than the averaged skewness of the given data (red line in Figure \ref{SM-fig_skewness_count}). As $p<0.05$, we rejected $H_{0}$ in support of significantly stronger lower tail dependence which was also consistent with the non-parametric analysis and model fitting results. On the other hand, Figure \ref{SM-fig_skewness_ff} demonstrated that averaged skewness was much higher than expected if the aphid first flight data came from a normal copula. As $p$ value (the fraction of surrogate-skewness which were greater than the averaged skewness of the given data - red line in Figure \ref{SM-fig_skewness_ff}) was less than $0.05$, we reject $H_{0}$ in support of upper tail dependence. Figure \ref{SM-fig_skewness_pns} showed that averaged skewness was lower than expected if the plankton data came from a normal copula. Our previous results on non-parametric analysis and model fitting also supported lower tail dependence. But, as $p>0.05$, we could not reject $H_{0}$ in that case. Finally, for methane-flux data, the averaged skewness was lower than expected if the data came from a normal copula (Figure \ref{SM-fig_skewness_methane}). As $p<0.05$, we rejected $H_{0}$ and that result was consistent with the non-parametric analysis and model selection results which also favored lower tail dependence.

@@

In order to study the effects of non-Normal copula structure on extinction risk for metapopulations with different dispersal rules, we examined the following generalized multiple patch model where local population dynamics at each patch were governed by the Lewton-Cohen model ($r=0$) or Ricker model ($r \neq 0$) as the case may be;
<!--did I write correct spelling for lewton model?-->

@@

$r$ is the liner growth rate and $K$ is the carrying capacity of the environment. $\epsilon$ is the environmental noise which might take stronger right tail (means stronger upper tail)  or stronger left tail (means stronger lower tail) dependence. We then considered each population $P(t+1)$ was updated depending upon $D$, the dispersal matrix, which could take either forms; one case when each population patch on a linear chain interacted with its two nearest neighbors (we called it local dispersion) or when any patch was allowed to interact with the rest of all (we called it as global dispersion). We simulated Eq. \ref{eq.extrisk_model} for $10000$ simulations and calculated extinction risk after $25$ years with different possibilities (for code see [mydispersal.R](https://github.com/sghosh89/BIVAN/blob/master/mydispersal.R)) and summarized our findings as follows;

@@

    * For LC model, as expected, when extinction risk calculated after $25$ years, was plotted against each varying $d$ (=degree of dispersion), we found that extinction risk was higher for data came from left tail dependent environmental fluctuations than right tail dependent one. [see first subplot of Fig. \ref{SM-fig_risk_local_disp} and Fig. \ref{SM-fig_risk_global_disp}]

@@

### Effect on Taylor's law

Spatial Taylor's law, which shows linear relationship between `log`(variance) and `log`(mean) of population densities of a species from different locations, is a well-known pattern found in ecology and related fields [@Taylor1961, @Taylor1988, @Cohen2013, @Doring2015, @Kuo2016, @Xu2016, @Reuman2017]. Here, we examined the nature of the slope and how it changed if population time series (taken as $50$ years) from different locations (say, $numloc=25$) had some certain type of copula structure (namely, \textbf{C}, \textbf{F} and \textbf{SC}) other than the $\textbf{N}$ormal copula. First, we simulated (N $\times$ n $\times$ nr) dimensional normal copula with covariance $\rho =0.7$, where $(N=50,n=25,nr=10)$ are the length of timeseries, number of sampling locations and number of replica for that copula, respectively. Then we generated Kendall-correlation preserving \textbf{C}layton, \textbf{F}rank and $\textbf{S}$urvival \textbf{C}layton surrogates with same dimension of \textbf{N}ormal. With these four different types of copula we computed `log`(spatial variance) and `log`(mean) and plotted in Fig.\ref{fig_spTL}(A-D). We also tested statistically that whether linearity, homoskedasticity is maintained in each copula dataset or not and plotted our results alongwith root mean square error, intercept and slope in Fig.\ref{fig_spTL}(E-I) (For code see [`copsurrognd.R`](https://github.com/sghosh89/BIVAN/blob/master/copsurrognd.R) and [`TaylorsLawTools.R`](https://github.com/sghosh89/BIVAN/blob/master/TaylorsLawTools.R)).
